\large How resource abundance and stochasticity affect animals' spatial needs

### 2023-07-25 writing/manuscript.Rmd
updated based on Sarah's comments

\noindent The amount of resources an organism is able to access is a strong determinant of its odds of survival and reproduction. Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], and even death [@foley_severe_2008; @berger_climate_2018]. Thus, many organisms will adapt their behaviors in response to changes in local resource abundance to ensure their needs are met. Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by reducing body heat [@schmidt_interplay_2020], or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020], but movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @wheat_migrate_2017; mammals: @singh_migration_2012], or moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019].

@@

\noindent Random variables are defined by specifying the **distribution** the variable is assumed to follow (e.g., Gaussian, Gamma, Poisson, Beroulli). Since the variable is random, it can take multiple possible values, each with different probabilities. The set or range of values which have non-zero probabilities in a distribution are referred to as the distribution's **support**.

There are many distributions we can assign to $R$, depending on how we quantify it. For instance, if $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow a distribution with support over all real numbers between 0 and infinity, which we can indicate with $R \in (0, \infty)$,^[We use round parentheses rather than square brackets (i.e., $[0, \infty]$) because neither 0 nor infinity are included in the distribution's support set.] such as a Gamma or log-normal distribution. If $R$ is a discrete variable, such as the number of prey in a location during a period of time, we can use a Poisson or negative binomial distribution. Alternatively, if we measure $R$ using the Normalized Difference Vegetation Index [**NDVI**, see @pettorelli_normalized_2011], we should use a distribution with support over the interval $[-1, 1]$, since NDVI can only take on values between -1 and 1 (extremes included).

@@

\noindent Since the exact value of $R$ at a given time and location is unknown, comparing the magnitude in $R$ between two locations or time periods requires a quantitative measure of what value we believe $R$ will take, on average. The **expectation** of a random variable is the value one can *expect* the random variable to take, on average, in the long term. Thus, we can let $\e(R)$ indicate the expectation of the random variable $R$. When the mean changes over time, such as in strogly seasonal regions, we explicitly indicate that $\e(R)$ changes over time by writing $\e(R)$ as a function of time, $t$: $\mu(t)$. We indicate the *estimated* average resource abundance over time by adding a caret on the symbol: $\hat \mu(t)$.
s
### Variance in resource abundance, $\var(R)$

\noindent In viewing resources as a random variable, we not only obtain a formal framework for describing $\e(R)$, but also the spread around $\e(R)$. A random variable's **variance** is a measure of its unpredictability (i.e., variability). We use the notation $\var(R)$ to indicate the variance in $R$ after accounting for changes in $\mu(t)$ and $\widehat{\sigma^2}(t)$ to indicate its estimate over time. Consequently, while $R$ may vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is (predictably) lower during some seasons and higher in others. Additionally, we distinguish between two different sources of variation: We refer to known or expected spatiotemporal differences or changes as **heterogeneity**, which we consider as a change in $\mu(t)$. In contrast, we use the term **stochasticity** for unpredictable and unexpected variation (i.e., due to $\sigma^2(t)$. For example, the location of (high-yield) berry bushes in a valley and the time at which they produce fruit may be heterogeneous, but it becomes predictable once a bear visits all patches in the valley. In contrast, whether next year will be a good or bad year for berries is stochastic because it depends on factors the bear cannot predict.

@@

\noindent In this paper, we simulate $R$ using a Gamma distribution with time-dependent mean $\mu(t)$ and time-dependent variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$. Although Gamma distributions are more often parameterized using parameters shape and scale ($k, \theta$) or shape and rate ($\alpha, \beta = 1/\theta$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations. However, note that $\e(R)$ and $\var(R)$ are not independent because the variance depends strongly on the mean (and vice-versa): As the mean approaches zero (from the positive side, which we write as $\e(R) \rightarrow0^+$) the variance also does: $\e(R) \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \var(R) = 0^+$. This assumption also holds biologically: In the example from above, a mean of zero kg of berries within the bear's habitat during winter implies the variance in berries is also zero.

@@

When $\e(R)$ is high, we expect animals' spatial requirements be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005]. However, as $\e(R)$ decreases, we expect animals' spatial requirements to increase nonlinearly, since low values of $\e(R)$ force animals to expand their home ranges [@lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019]. It is unclear when animals switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between these types of movement is important for quantifying the effect of resource abundance on animal's spatial needs and when an animal may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020]. Sill, large-scale changes in movement behavior (such as species-wide shifts to dispersal, migration, or nomadism) are more likely to occur over evolutionary timescales than over an animal's lifespan. For instance, larger home ranges requires greater vagility, which, in turn, is facilitated by morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].

@@

```{r multipanel-constant-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with constant variance (after accounting for changes in the mean). (a.) Resources ($R$, brown-green fill) vary over time and space with a constant the variance (dot color). (b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and constant variance $\\var(R) = \\sigma^2$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/habitat-examples-constant-variance.png')
```

Consider the scenario where the average in resource abundance, $\e(R)$, changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (see figure \@ref(fig:multipanel-constant-var)a). We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b). Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat. However, if $\mu(t)$ changes repetitively and regularly over time (e.g. peaks each spring and is lowest in winter), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017], but it will not be able to predict *where* $R$ is high if $\e(R)$ does not vary predictably over space.

@@

Similar to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on animals' spatial needs. When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect spatial needs to increase progressively faster (figure \@ref(fig:s2-hyp)) because (1) as $\var(R)$ increases the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], and (3) the added movement required to search for food increases animals' energetic requirements. If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage or food caching). If changes in $\sigma^2(t)$ are sufficiently predictable, animals may learn to anticipate and prepare for periods of greater stochasticity by pre-emptively caching food or migrating, or relying on alternative food sources during stochastic events (e.g., after the arrival of competitors).

```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with changing variance (even after accounting for changes in the mean). (a.) Although resource abundance ($R$, brown-green fill) varies over time and space, the variance (dot color) is lowest at the beginning and end of the observational period and highest when $\\e(R)$ peaks. (b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and variance $\\sigma^2(t)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
```

@@

@rizzuto_forage_2021 found that the spatial needs of snowshoe hares (*Lepus americanus*) increased with average carbon to nitrogen ratio (C:N, a measure of N scarcity) in lowbush blueberry as well as the coefficient of variation (the standard deviation over the mean, $\frac{\mu(t)}{\sqrt{\sigma^2(t)}}$) in C:N. However, it is hard to determine how $\sigma^2(t)$ affected the hares' spatial needs since the coefficient of variation is a function of both $\mu(t)$ and $\sigma^2(t)$, so the effects of the two variables are confounded. Similarly, @mueller_how_2011 suggested that ecosystem variability (measured as spatial semivariance in NDVI) causes ungulates to move more and be more likely to be nomadic, but it is hard to determine whether the higher semivariances in NDVI are due to an increase in $\var(R)$ or simply spatial changes in $\e(R)$ (e.g., increased seasonality, heterogeneity, or continentality). Likewise, @herfindal_prey_2005 found that the home ranges Canadian lynx (*Lynx canadensis*) decreased with seasonality, but this may be an artifact of a lower $\mu(t)$ in highly seasonal environments, rather than stochasticity in $R$. A subsequent analysis by @nilsen_can_2005 of the lynx data of @herfindal_prey_2005 along with data on wolverines (*Gulo gulo*), bobcats (*Lynx rufus*), and coyotes (*Canis latrans*) found that carnivores respond differently to seasonality, which suggests that different species may have different perceptions of the changes in $R$ over the seasons. Collectively, these studies provide support towards the effects of both $\e(R)$ and $\var(R)$ on animals' spatial needs, but the lack of a unifying framework across these studies makes it challenging to synthesize their findings. Thus, it is important to disentangle the effects of spatiotemporal heterogeneity (i.e., changes in $\e(R)$, such as seasonal trends in $\mu(t)$) and stochasticity ($\var(R)$). The need to account for changes in both $\e(R)$ and $\var(R)$ is compounded by recent changes in climate, which expose species to novel situations and increasingly common and extreme stochastic events [@ipcc_global_2018; @noonan_situ_2018; @yao_emergence_2022].

@nilsen_can_2005 also show that the effect of $\e(R)$ on the home-range size of wolves and fishers (*Martes pennanti*) depends on seasonality (and vice-versa): Animals in habitats with high and low seasonality had opposite responses to $\e(R)$. Therefore, it may not be sufficient to account for the effects of $\e(R)$ and $\var(R)$; models may also need a term for interaction between the two. This may be because animals respond strongly to $\sigma^2(t)$ when $\mu(t)$ is low, but the they may not respond if $\mu(t)$ is sufficiently high. However, it does not explain why animals may respond to $\e(R)$ in opposite ways in different habitats. Instead, we suspect this inconsistency is because @nilsen_can_2005 do not distinguish between changes in $\mu(t)$ and $\sigma^2(t)$ between seasons, so the effects of the two are confounded in the seasonality term. Distinguishing between changes in $\e(R)$ and $\var(R)$ allows one to separate the two seasonal cycles and produce more consistent results.

@@

```{r 5-5-sims, fig.cap="Simulated home range sizes, $H$, of an animal living in habitats where the mean and variance in resources are constant, linearly increasing, cyclical, drifting, or erratic over time. Note how $H$ decreases nonlinearly as $\\mu(t)$ increases and increases nonlinearly as $\\sigma^2(t)$ increases. Additionally, the variance in $H$ is higher when $\\mu(t)$ is lower or $\\sigma^2(t)$ is higher, and changes in $\\sigma^2(t)$ have greater impacts when $\\mu(t)$ is low.\\newline{}Simulations were run such that animals followed the same $2^{10}=1024$ tracks at each time point starting from the location $\\langle 0, 0\\rangle$ until they reach satiety, after which they were allowed to return to $\\langle 0, 0\\rangle$ over the same amount of time required to reach satiety. The animal's movement and utilization distribution were then estimated using an Ornstein-Uhlenbeck Foraging (OUF) model and Autocorrelated Kernel Density Estimation (AKDE), respectively, via the $\\tt{ctmm}$ package (Fleming and Calabrese 2021). $H$ was estimated using the UD's 0.95 quantile, but the effect is true for all quantiles since the UD was estimated via AKDE. Additional information is provided in Appendix 1.", out.width='100%'}
knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png')
```

@@

Although the temporal trends in figure \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two surprisingly simple relationships emerge when home-range size is shown as a function of either E(R) or V(R), rather than time. Both $\e(R)$ and $\var(R)$ affect spatial needs precisely as we hypothesized (figure \@ref(fig:5-5-reg)).<!--, although we found that the effect of $\var(R)$ is closer linear than we expected (it is mostly sub-linear as long as $H$ is sufficiently far from zero). We ensured this was not due to a low maximum $\var(R)$ (as in the left-hand side of figure \@ref(fig:s2-hyp)) by simulating within the range $\sigma^2(t) \in [2\,500, 250\,000] \implies \sigma \in [50, 500]$, since the range of $\e(R)$ was $\mu(t) \in [100, 200]$.--> We believe the approximately linear effect of $\var(R)$ on $H$ is mainly due to two factors. Firstly, $H$ scales superlinearly with the standard diviation in $R$ (as $\sqrt{\var(R)}$ is on the same scale as $\e(R)$; not shown), which would necessarily imply that $H$ increases *sub*linearly with $\var(R)$. Secondly, the animal's movement was simulated to be range resident within an homogeneous environment. Since real-world landscapes are spatially heterogeneous in both $\e(R)$ and $\var(R)$, animals may choose to shift their home range to a new area following a stochastic event that caused $\mu(t)$ to decrease, such as a fire or a flood. Thus, we hypothesize that our simulations may be underestimating animals' responses to spatiotemporal changes in $\e(R)$ and $\var(R)$, since the simulated animals cannot select for areas of higher $\e(R)$ or lower $\var(R)$. Additionally, the simulations also ignore the how competition, movement costs, predation, and other factors may affect the effects of $\e(R)$ and $\var(R)$ on $H$. While these are definitely important effects that should be accounted for [@jetz_scaling_2004], doing so was beyond the scope of this paper.

```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements. The relationships were estimated using a Generalized Additive Model for Location and Scale (Rigby and Stasinopoulos 2005; Stasinopoulos and Rigby 2007) with a Gamma family of distributions. The model accounted for the effects of $\\e(R)$ and $\\var(R)$ and the interaction effects between the two on both $\\e(H)$ and $\\var(H)$. Note the nonlinear decrease in $H$ as $\\e(R)$ increases and the approximately nonlinear increase in $H$ as $\\var(R)$ increases. Additionally, note how the variance in spatial needs increases with mean spatial needs.", out.width='100%'}
knitr::include_graphics('../figures/simulation-regression-plots.png')
```

# Applying the framework: The effects of E(R) and V(R) on a lowland tapir

>>>>

\noindent The amount of resources an organism is able to access is a strong determinant of its odds of survival and reproduction. Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], and even death [@foley_severe_2008; @berger_climate_2018]. Thus, many organisms will adapt their behaviors in response to changes in local resource abundance to ensure their needs are met. Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by lowering body heat [@schmidt_interplay_2020], or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020]. However, movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], or dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @wheat_migrate_2017; mammals: @singh_migration_2012].

@@

\noindent Random variables are defined by specifying the **distribution** the variable is assumed to follow (e.g., Gaussian, Gamma, Poisson, Beroulli). Since the variable is random, it can take multiple possible values, each with different probabilities. The set or range of values which have non-zero probabilities in a distribution is referred to as the distribution's **support**.

There are many distributions we can assign to $R$, depending on how we quantify it. For instance, if $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow a distribution with support over all real numbers between 0 and infinity, which we can indicate with $R \in (0, \infty)$,^[We use round parentheses rather than square brackets (i.e., $[0, \infty]$) because neither 0 nor infinity are included in the distribution's support set.] such as a Gamma or log-normal distribution. If $R$ is a discrete integer variable, such as the number of prey in a location during a period of time, we can use a Poisson or negative binomial distribution. Alternatively, if we measure $R$ using the Normalized Difference Vegetation Index [**NDVI**, see @pettorelli_normalized_2011], we should use a distribution with support over the interval $[-1, 1]$, since NDVI can only take on values between -1 and 1 (extremes included). However, since there is no commonly used distribution with that support, the best option is to rescale NDVI to (0, 1) and use a beta distribution (see the section below on applying this framework).

@@

\noindent Since the exact value of $R$ at a given time and location is unknown, comparing the magnitude in $R$ between two locations or time periods requires a quantitative measure of what value we believe $R$ will take, on average. The **expectation** of a random variable is the value one can *expect* the random variable to take, on average, in the long term. Thus, we can let $\e(R)$ indicate the expectation of the random variable $R$. When the mean changes over time, such as in strongly seasonal regions, we explicitly indicate that $\e(R)$ changes over time by writing $\e(R)$ as a function of time, $t$: $\e(R|t) = \mu(t)$. We indicate the *estimated* average resource abundance over time by adding a caret on the symbol: $\hat \mu(t)$.

### Variance in resource abundance, $\var(R)$

\noindent In viewing resources as a random variable, we not only obtain a formal framework for describing $\e(R)$, but also the spread around $\e(R)$. A random variable's **variance** is a measure of its unpredictability (i.e., variability). We use the notation $\var(R)$ to indicate the variance in $R$ after accounting for changes in $\mu(t)$, and we use $\sigma^2(t)$ to indicate its function over time (with estimate $\widehat{\sigma^2}(t)$). Consequently, while $R$ may vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is (predictably) lower during some seasons and higher in others. Additionally, we distinguish between two different sources of variation. We refer to known or expected spatiotemporal differences or changes as **heterogeneity**, which we consider as a change in $\mu(t)$. In contrast, we use the term **stochasticity** for unpredictable and unexpected variation (i.e., due to $\sigma^2(t)$. For example, the location of (high-yield) berry bushes in a valley and the time at which they produce fruit may be heterogeneous, but it becomes predictable once a bear visits all patches in the valley. In contrast, whether next year will be a good or bad year for berries is stochastic because it depends on factors the bear cannot predict.

@@

\noindent In this paper, we simulate $R$ using a Gamma distribution with time-dependent mean $\mu(t)$ and time-dependent variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$. Although Gamma distributions are more often parameterized using parameters shape and scale ($k > 0$ and $\theta > 0$) or shape and rate ($\alpha > 0$ and $\beta = 1/\theta > 0$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations. However, note that $\e(R)$ and $\var(R)$ are not independent because the variance depends strongly on the mean (and vice-versa). As the mean approaches zero (from the positive side, which we write as $\e(R) \rightarrow0^+$) the variance also does: $\e(R) \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \var(R) = 0^+$. This assumption also holds biologically; in the example from above, a mean of zero kg of berries within the bear's habitat during winter implies the variance in berries is also zero.

@@

When $\e(R)$ is high, we expect animals' spatial requirements be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005]. However, as $\e(R)$ decreases, we expect animals' spatial requirements to increase nonlinearly, since low values of $\e(R)$ force animals to expand their home ranges [@lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019]. It is unclear when animals switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between these types of movement is important for quantifying the effect of resource abundance on animal's spatial needs and when an animal may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020]. Still, large-scale changes in movement behavior (such as species-wide shifts to dispersal, migration, or nomadism) are more likely to occur over evolutionary timescales than over an animal's lifespan. For instance, larger home ranges requires greater vagility, which, in turn, is facilitated by morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].

@@

```{r multipanel-constant-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with constant variance (after accounting for changes in the mean). (a.) Resources ($R$, brown-green fill) vary over time and space with a constant the variance (dot color). (b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and constant variance $\\var(R) = \\sigma^2$. The points in the central panel indicate the ralizations of $R$ in panel a.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/habitat-examples-constant-variance.png')
```

Consider the scenario where the average in resource abundance, $\e(R)$, changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (see figure \@ref(fig:multipanel-constant-var)a). We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b). Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat. However, if $\mu(t)$ changes repetitively and regularly over time (e.g. peaks each spring and is lowest in winter), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017], but it will not be able to predict *where* $R$ is high if $\e(R)$ does not vary predictably over space.

@@

Similar to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on animals' spatial needs. When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect spatial needs to increase progressively faster (figure \@ref(fig:s2-hyp)) because: (1) as $\var(R)$ increases the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], and (3) the added movement required to search for food increases animals' energetic requirements. If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage or food caching). If changes in $\sigma^2(t)$ are sufficiently predictable, animals may learn to anticipate and prepare for periods of greater stochasticity by pre-emptively caching food or migrating, or relying on alternative food sources during stochastic events (e.g., after the arrival of competitors).

```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with changing variance (even after accounting for changes in the mean). (a.) Although resource abundance ($R$, brown-green fill) varies over time and space, the variance (dot color) is lowest at the beginning and end of the observational period and highest when $\\e(R)$ peaks. (b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and variance $\\sigma^2(t)$. The points in the central panel indicate the ralizations of $R$ in panel a.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
```

@@

@rizzuto_forage_2021 found that the spatial needs of snowshoe hares (*Lepus americanus*) increased with average carbon to nitrogen ratio (C:N, a measure of N scarcity) in lowbush blueberry as well as the coefficient of variation (the standard deviation over the mean, $\frac{\sqrt{\sigma^2(t)}}{\mu(t)}$) in C:N. However, it is hard to determine how $\sigma^2(t)$ affected the hares' spatial needs since the coefficient of variation is a function of both $\mu(t)$ and $\sigma^2(t)$, so the effects of the two variables are confounded. Similarly, @mueller_how_2011 suggested that ecosystem variability (measured as spatial semivariance in NDVI) causes ungulates to move more and be more likely to be nomadic, but it is hard to determine whether the higher semivariances in NDVI are due to an increase in $\var(R)$ or simply spatial changes in $\e(R)$ (e.g., increased seasonality, heterogeneity, or continentality). Likewise, @herfindal_prey_2005 found that the home ranges Canadian lynx (*Lynx canadensis*) decreased with seasonality, but this may be an artifact of a lower $\mu(t)$ in highly seasonal environments, rather than stochasticity in $R$. A subsequent analysis by @nilsen_can_2005 of the lynx data of @herfindal_prey_2005 along with data on wolverines (*Gulo gulo*), bobcats (*Lynx rufus*), and coyotes (*Canis latrans*) found that carnivores respond differently to seasonality, which suggests that different species may have different perceptions of the changes in $R$ over the seasons. Collectively, these studies provide support towards the effects of both $\e(R)$ and $\var(R)$ on animals' spatial needs, but the lack of a unifying framework across these studies makes it challenging to synthesize their findings. Thus, it is important to disentangle the effects of spatiotemporal heterogeneity (i.e., changes in $\e(R)$, such as seasonal trends in $\mu(t)$) and stochasticity ($\var(R)$). The need to account for changes in both $\e(R)$ and $\var(R)$ is compounded by recent changes in climate, which expose species to novel situations and increasingly common and extreme stochastic events [@intergovernmental_panel_on_climate_change_climate_2023; @noonan_situ_2018; @yao_emergence_2022].

@nilsen_can_2005 also show that the effect of $\e(R)$ on the home-range size of wolves and fishers (*Martes pennanti*) depends on seasonality (and vice-versa), since animals in habitats with high and low seasonality had opposite responses to $\e(R)$. Therefore, it may not be sufficient to account for the effects of $\e(R)$ and $\var(R)$; models may also need a term for interaction between the two. This may be because animals respond strongly to $\sigma^2(t)$ when $\mu(t)$ is low, but they may not respond if $\mu(t)$ is sufficiently high. However, it does not explain why animals may respond to $\e(R)$ in opposite ways in different habitats. Instead, we suspect this inconsistency is because @nilsen_can_2005 do not distinguish between changes in $\mu(t)$ and $\sigma^2(t)$ between seasons, so the effects of the two are confounded in the seasonality term. Distinguishing between changes in $\e(R)$ and $\var(R)$ allows one to separate the two seasonal cycles and produce more consistent results.

@@

```{r 5-5-sims, fig.cap="Simulated home range sizes, $H$, of an animal living in habitats where the mean and variance in resources are constant, linearly increasing, cyclical, drifting, or erratic over time. Note how $H$ decreases nonlinearly as $\\mu(t)$ increases and increases nonlinearly as $\\sigma^2(t)$ increases. Additionally, the variance in $H$ is higher when $\\mu(t)$ is lower or $\\sigma^2(t)$ is higher, and changes in $\\sigma^2(t)$ have greater impacts when $\\mu(t)$ is low.\\newline{}Simulations were run such that animals followed the same $2^{10}=1024$ tracks at each time point starting from the location $\\langle 0, 0\\rangle$ until they reach satiety, after which they were allowed to return to $\\langle 0, 0\\rangle$ over the same amount of time required to reach satiety. The animal's movement and utilization distribution were then estimated using an Ornstein-Uhlenbeck Foraging (OUF) model and Autocorrelated Kernel Density Estimation (AKDE), respectively, via the $\\tt{ctmm}$ package (Fleming and Calabrese 2021) for $\\tt{R}$ (R Core Team 2022). $H$ was estimated using the UD's 0.95 quantile, but the effect is true for all quantiles since the UD was estimated via AKDE. Additional information is provided in Appendix 1.", out.width='100%'}
knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png')
```

@@

Although the temporal trends in figure \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two surprisingly simple relationships emerge when home-range size is shown as a function of either $\e(R)$ or $\var(R)$, rather than time. Both $\e(R)$ and $\var(R)$ affect spatial needs precisely as we hypothesized (figure \@ref(fig:5-5-reg)).<!--, although we found that the effect of $\var(R)$ is closer linear than we expected (it is mostly sub-linear as long as $H$ is sufficiently far from zero). We ensured this was not due to a low maximum $\var(R)$ (as in the left-hand side of figure \@ref(fig:s2-hyp)) by simulating within the range $\sigma^2(t) \in [2\,500, 250\,000] \implies \sigma \in [50, 500]$, since the range of $\e(R)$ was $\mu(t) \in [100, 200]$.--> We believe the approximately linear effect of $\var(R)$ on $H$ is mainly due to two factors. Firstly, $H$ scales superlinearly with the standard diviation in $R$ (as $\sqrt{\var(R)}$ is on the same scale as $\e(R)$; not shown), which would necessarily imply that $H$ increases *sub*linearly with $\var(R)$. Secondly, the animal's movement was simulated to be range resident within an homogeneous environment. Since real-world landscapes are spatially heterogeneous in both $\e(R)$ and $\var(R)$, animals may choose to shift their home range to a new area following a stochastic event that caused $\mu(t)$ to decrease, such as a fire or a flood. Thus, we hypothesize that our simulations may be underestimating animals' responses to spatiotemporal changes in $\e(R)$ and $\var(R)$, since the simulated animals cannot select for areas of higher $\e(R)$ or lower $\var(R)$. Additionally, the simulations also ignore how competition, movement costs, predation, and other factors may affect the effects of $\e(R)$ and $\var(R)$ on $H$. While these are definitely important effects that should be accounted for [@jetz_scaling_2004], doing so was beyond the scope of this paper.

```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements. The relationships were estimated using a Generalized Additive Model for Location and Scale (Rigby and Stasinopoulos 2005; Stasinopoulos and Rigby 2007) with a Gamma family of distributions. The model accounted for the effects of $\\e(R)$ and $\\var(R)$ and the interaction effects between the two on both $\\e(H)$ and $\\var(H)$. Note the nonlinear decrease in $H$ as $\\e(R)$ increases and the increase in $H$ as $\\var(R)$ increases. Additionally, note how the variance in spatial needs increases with mean spatial needs.", out.width='100%'}
knitr::include_graphics('../figures/simulation-regression-plots.png')
```

# Applying the framework: The effects of $\e(R)$ and $\var(R)$ on a lowland tapir
